Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-04-28 curvature.tex
Some revisions to the size distribution of subtrees.

\begin{lem}
	\label{lem:compute_degree}
	For a tree $T$ with $n$ leaves,
	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
	$$\size{N(T,u)} = \begin{cases}
		2n - x - 5 &\text{if depth($u$) $ > 1$, } \\
		2n - x - 3 &\text{if depth($u$) $ = 1$ } \\
		0 &\text{if depth($u$) $ \le 0$ } \\
	\end{cases}.$$
\end{lem}
\begin{proof}
	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
	Call these $U$ and $V$, respectively.
	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (move type (iii)) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (move type (iv)), a contradiction.
	Thus the claim holds.
\end{proof}

@@

In our final two lemmas of this section, we consider the distribution of trees in a neighborhood with respect to the size of the subtrees moved by SPR moves.
First, we bound the number of neighbors of a tree that are induced by SPR moves involving a subtree with three or more leaves.
In particular, we show that the majority of neighbors are induced by small subtree moves (one or two leaves).
Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR move depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} SPR move is one that moves a small subtree.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 3$ leaves.
	Let $N_S(T)$ be the neighbors of $T$ induced by small SPR moves. 
	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
	\label{lem:spr_size}
\end{lem}
\begin{proof}
	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	Each subtree at a depth $ > 1$ from the root induces $2n - x - 5$ SPR neighbors, where $x$ is the number of nodes in the subtree, while the children of the root induce $2n - x - 3$ neighbors.
	This implies that the number of small SPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.
	There are thus two extremes for the distribution of rSPR moves induced by small subtrees.
	A ladder tree has the smallest proportion of small subtrees.
	One 1-leaf subtree in a ladder tree is a child of the root, but this is counterbalanced by the fact that any other tree has at least one more subtree with 2 leaves.
	A balanced tree has the greatest proportion of small subtrees.
	Thus, a balanced tree has the greatest number of small SPR moves.

	We first bound the number of small subtree moves that can be applied to a ladder tree.
	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.
	One of the 1-leaf subtrees is a child of the root.
	A 2-leaf subtree contains three nodes.
	Therefore, a ladder tree induces $n (2n - 1 - 5) + 2 + (2n - 3 - 5) = 2n^2 - 4n - 6$ small subtree moves.

>>>>

\todo[inline]{Please define x in the text of the following lemma.}
\begin{lem}
	\label{lem:compute_degree}
	For a tree $T$ with $n$ leaves,
	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
	$$\size{N(T,u)} = \begin{cases}
		2n - x - 5 &\text{if depth($u$) $ > 1$, } \\
		2n - x - 3 &\text{if depth($u$) $ = 1$ } \\
		0 &\text{if depth($u$) $ \le 0$ } \\
	\end{cases}.$$
\end{lem}
\begin{proof}
	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
	Call these $U$ and $V$, respectively.
	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (move type (iii)) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (move type (iv)), a contradiction.
	Thus the claim holds.
\end{proof}

@@

In our final two lemmas of this section, we consider the distribution of sizes of subtrees moved by SPR operations from a given tree.
First, we bound the number of neighbors of a tree that are induced by SPR operations involving a subtree with three or more leaves.
In particular, we show that the majority of neighbors are induced by moving small subtrees (one or two leaves).
Along with our observation from Lemma~\ref{lem:degree_change} that the magnitude of the degree change induced by an SPR operation depends on the subtree size and height change, this implies that most SPR neighbors have a similar degree.
	We say that a subtree is \emph{small} if it has two leaves or one leaf, and that a \emph{small} SPR move is one that moves a small subtree.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 3$ leaves.
	Let $N_S(T)$ be the neighbors of $T$ induced by small SPR moves.
	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
	\label{lem:spr_size}
\end{lem}
\begin{proof}
	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ SPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
	This implies that the number of small SPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.

	A ladder tree has the smallest proportion of small subtrees, so we first bound the number of small subtree moves that can be applied to a ladder tree.
	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.
	One of the 1-leaf subtrees is a child of the root.
	A 2-leaf subtree contains three nodes.
	Therefore, a ladder tree induces $n (2n - 1 - 5) + 2 + (2n - 3 - 5) = 2n^2 - 4n - 6$ small subtree moves.
